---
title: |
  A slice tour for finding hollowness in high-dimensional data
type: Short Technical Note
author:
  - name: Ursula Laa
    affil: a, b
    email: ursula.laa@monash.edu
  - name: Dianne Cook
    affil: b
    email: dicook@monash.edu
affiliation:
  - num: a
    address: |
      School of Physics and Astronomy, Monash University
  - num: b
    address: |
      Department of Econometrics and Business Statistics, Monash University
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: |
  Projections and sectioning are complementary methods for the visualisation of
  high-dimensional data. In this note we describe slicing in the orthogonal
  space of projections obtained when running a tour, thus presenting the viewer
  with an interpolated sequence of projected sections. The method has been
  implemented in R as an extension to the tourr package, and can be used to
  explore concave structures in multivariate distributions.
keywords: |
  data visualisation; grand tour; sectioning
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
output: rticles::tf_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

\doublespacing
# Introduction

<!-- Background -->
Modern data is typically high-dimensional and thus data visualisation often relies on some form of dimension reduction. This entails looking at linear projections of the data, or nonlinear if one considers techniques like multidimensional scaling (MDS) or t-Distributed Stochastic Neighbor Embedding (t-SNE). For our purposes we focus on linear projections. Interactive and dynamic displays can provide information beyond what can be achieved in a static display. For example, the grand tour [@As85; @BCAH05] shows a smooth sequence of interpolated lower dimensional projections and allows the viewer to extrapolate from the low-dimensional shapes to the multidimensional distribution. It is particularly useful for detecting clusters, outliers and non-linear dependence. 

<!-- The disadvantage of projection: opacity, occlusion, even with pseudo-transparency from scatterplots;  motivation -->
A major limitation of projections is their opacity, even if mitigated somewhat through the pseudo-transparancy present in sparse scatterplots. As a result certain features of distributions may easily be hidden in projections, especially in the case of concave structures. Considering the example of a simple geometric shape such as a hypershere, we understand that projections cannot distinguish between a full or a hollow sphere. Similarly we can think of small scale structures hidden in the center of a multivariate distribution. Again projections will not help in findig such features in a distribution. Other examples where the use of projections is limited include the visualisation of non-linear boundaries between groups, that will be obscured in each projection.

<!-- Our approach: introduce sections; many different ways to slice, refer to linked brushing as slicing whcih mirrors prosection, but here we will explain/utilise a simple slicing on the projection where point in orthogonal space are faded. -->
We can overcome these limitations when adding section information, see e.g. @prosection. For our example of the hypersphere, we can clearely identify a hollow sphere when looking at a slice throught the distribution, where all points will lie on a circle in a 2d slice. Note that there are many ways to define a slice through high-dimensional data. In practice, slicing is typically facilitated through linked brushing (e.g. @ggobiCSDA2003, @JSSv081i05), where the user is selecting a region in a subspace of the dataspace that is highlighted in a projection of the data points. These systems are interactive, and rely on the user to redefine the section many times to explore the data in this way. In this note we introduce a simple slicing method for projections where points in the orthogonal space are faded. Combining it with the grand tour results in a new dynamic display that can be used to systematically search for features hidden in sections.


<!-- paper structure --> 
The method and its implementation in R are described in Section XXX, and in Section XXX. We then apply it to several examples in Section XXX. We consider different high-dimensional geometric shapes and XXX, showing how concave or occluded structures can be visualised and explored with a slice tour. Possible extensions are discussed in Section XXX.

# Method

The basis underlying our approach is the grand tour. We first briefly review the relevant concepts. Building on these ideas we introduce a method for displaying slices of parameter space that are defined for each projection matrix.

## Tour review

A tour is providing a continuous animation, showing a sequence of $d$-dimensional (typically $d=1$ or $2$) projections from $p$-dimensional parameter space. The tour combines basis selection with basis interpolation. In a grand tour the basis selection is random (each new basis is chosen from all possible projections), and typically geodesic interpolation is used to generate smooth transitions between viewing planes. Different basis selection methods as well as the geodesic interpolation are implemented in the `tourr` package [@tourr] in R [@rref], which also provides several display functions for viewing the tour.

The notation used in the following for the description of each projection is denoting $\mathbf{X}$ the $n\times p$ dimensional data matrix ($n$ observations in $p$ dimensions) and $\mathbf{A}$ an orthonormal $p \times d$ projection matrix. The $d$-dimensional projection of the data is thus given by $\mathbf{Y} = \mathbf{X}\cdot\mathbf{A}$, giving the $n\times d$ dimensional projected data matrix to be plotted in each frame of the tour display.

In the following we are interested in the orthogonal distance of scatter points from an anchored projection plane, thus $d=2$. 

## Slicing in the orthogonal space

### Distance from the origin

In addition to the projected data $\mathbf{Y}$ we calculate for each projection the orthogonal distance of the data points from the current projection plane centered at the origin, $v$. The orthogonal component is given by
\begin{equation}
x_i' = x_i - (x_i\cdot a_1) a_1 - (x_i\cdot a_2) a_2
\end{equation}
with $x_i, i=1,...,n$ an $p$-dimensional observation in $\mathbf{X}$ and $a_i, i=1,2$ denoting the columns of the projection matrix, $\mathbf{A}=(a_1, a_2)$.
Calculating the euclidean norm of $x'$ is straightforward thanks to the orthonormality of $\mathbf{A}$, and gives the components of the required distance vector.
\begin{equation}
\tilde{v}_i^2 = ||x_i'||^2 = x_i^2 - (x_i\cdot a_1)^2 - (x_i\cdot a_2)^2.
\end{equation}
We also see that this is trivially generalizing for orthonormal projection matrices with $d>2$.

The distance vector can then be used to display a slice tour by highlighting points for which the orthogonal distance is smaller than a selected cutoff value $h$. Notice that in 3d, where for each plane we have a single orthogonal direction, this defines a flat slice of hight $2*h$, matching the common conception of a slice. When going beyond 3d, the codimension of the plane is $>1$ and using the euclidean distance results in rotation invariant slicing in the orthogonal space. As a consequence the "slices" are now spherical in the orthogonal subspace, with radius $h$.

```{r renderFunction, echo=FALSE, eval=TRUE, warning=FALSE, fig.height=1.5, fig.width = 5.5, fig.cap="Projected slices through 4d geometric shapes. On the left a full (hollow) hypersphere (a, b), on the right a full (hollow) hypercube (c, d). Points inside the slice are shown as black bullets, points outside the slice are shown as grey dots. We can clearly distinguish the full from the hollow objects based on the slice display.\\label{fig:renderFunction}"}
library(tidyverse)

# slight modification of the function in tourr::display_slice
render_data <- function(data, proj, h, anchor) {
    # Render projected points
    x <- data %*% proj
    d <- tourr::anchored_orthogonal_distance(proj, data, anchor)
    pch <- rep(46, nrow(x))
    pch[d < h] <- 20
    col <- rep("grey", nrow(x))
    col[d < h] <- "black"
    ggplot() +aes(x[,1], x[,2]) + geom_point(pch = pch, color=col) + theme_void() + coord_fixed()
}

anchor <- rep(0, 4)
plane <- tourr::basis_init(4, 2)
nPt <- 5000

set.seed(3008)

d1 <- geozoo::sphere.solid.random(4, nPt)$points
d2 <- geozoo::sphere.hollow(4, nPt)$points
d3 <- geozoo::cube.solid.random(4)$points[1:nPt,] %>% scale(scale = FALSE)
d4 <- geozoo::cube.face(4)$points[1:nPt,] %>% scale(scale = FALSE)

a <- render_data(d1, plane, 0.2, anchor) + labs(title="a")
b <- render_data(d2, plane, 0.2, anchor) + labs(title="b")
c <- render_data(d3, plane, 0.2, anchor) + labs(title="c")
d <- render_data(d4, plane, 0.2, anchor) + labs(title="d")

library(gridExtra)
grid.arrange(a, b, c, d, nrow=1)

```


### Slice thickness

Choosing the slice thickness is a compromise between what feature size can be revealed and how many points fall inside a thin slice. As dimensionality $p$ increases, the relative number of points that are inside a slice of fixed thickness $h$ will decrease. The exact relation depends heavily on the distribution of the data points. Here we present expressions for special cases to get an idea of this scaling. For practical purposes we specify a volume $\epsilon$ and then use as a cutoff $h = \epsilon^{1/(p-2)}$. XXX not yet implemented that way, based on thickness estimates.

We consider two types of rotation invariant distributions for our estimates: a hypersphere in $p$ dimensions and a $p$ dimensional normal distribution without correlation.

Let us consider points uniformly distributed in a $p$-dimensional hypersphere. The fraction of points inside a slice of thickness $2*h$ can therefore be estimated as the relative volume of the slice compared to the volume of the full hypersphere,
\begin{equation}
V_{rel} = \frac{1}{2} \frac{h^{p-2}}{r^p} (p r^2 - (p-2) h^2) \approx \frac{1}{2} (\frac{h}{R})^{p-2},
\end{equation}
where the approximation is valid when $h \ll r$, as is typically expected to be the case. Thus if we want to keep the relative number of points (i.e. $V_{rel}$) approximately constant for slices in different dimensions $p$, we calculate $h$ from a volume parameter as $h = \epsilon^{1/(p-2)}$. XXXX change code to this? currently $\epsilon^{1/p}$ is implemented...

XXXX need to redo integration of multivariate normal...
Considering now slicing through a multivariate normal, we can estimate the number of points in a slice by integrating the probability distribution with boundaries $\pm h$ in $p-2$. In this case the relative density scales with the error function as
\begin{equation}
\rho_{rel} = \pi^{(2-p)/2} \mathrm{erf}((h/\sqrt 2 \sigma)) ^{p-2}.
\end{equation}

XXX two plots: vrel/rhorel scaling with h for n for p = 3, 4, 5, 10, and h scaling with p for vrel = 0.001, 0.01, 0.05 and 0.1

### Anchoring

The above definition implicitly assumes that we slice the distribution using the origin as a reference point. This is generally a good starting point assuming the data is centered before applying the slice tour. Alternative reference points can however provide additional insight into the concave structure of the distribution. We can also use the mean in each parameter as an anchor point if the data is not centered.

We therefore generalize the distance vector to measure the anchored orthogonal distance, i.e. distance with respect to the anchor $\alpha$ used as a reference point.

In general $\alpha$ can be any point in the $n$-dimensional parameter space, but we are only interested in the orthogonal component, thus we use $\alpha' = \alpha - (\alpha\cdot a_1) a_1 - (\alpha\cdot a_2 )a_2$ to define the generalized measure of orthogonal distance as
\begin{equation}
v_i^2 = ||x_i' - \alpha'||^2 = x_i'^2 + \alpha'^2 - 2 x_i'\cdot\alpha'
\end{equation}

where the cross term can be expressed as

\begin{equation}
x_i'\cdot\alpha' = x_i\cdot\alpha - (\alpha\cdot a_1) (x_i\cdot a_1) - (\alpha\cdot a_2) (x_i\cdot a_2).
\end{equation}

Using the generalized distance measure with a cutoff volume $\epsilon$ then corresponds to moving a slice of fixed thickness to be centered around the anchor point $\alpha$ in the space orthogonal to the current projection plane. In three dimensions this simply corresponds to moving up or down along the orthogonal direction.



# Implementation

The slice tour has been implemented in R [@rref] as a new display method `display_slice` in the `tourr` package [@tourr]. In addition to usual parameters the user can choose the volume parameter $\epsilon$ by setting `eps` and, if required, the anchor point $\alpha$ is set by the `anchor` argument. By default `eps=0.001` and `anchor=NULL`, resulting in slicing through the origin. In addition the user can select the marker symbols for point inside and outside the slice. By default points in the slice are highlighted as `pch_slice  = 20` and `pch_other = 46`, i.e. plotting a bullet for points inside the slice, and a dot for points outside.

```{r codeExampel, echo=TRUE, eval=FALSE}
library(tourr)
# use geozoo to generate points on a hollow 3d sphere
sphere3 <- geozoo::sphere.hollow(3)$points
colnames(sphere3) <- c("x1", "x2", "x3") # naming variables
# slice tour animation with default settings
animate_slice(sphere3)
# trying an off-center anchor point, thicker slice, and
# we use pch=26 to hide points outside the slice
anchor3 <- rep(0.7, 3)
animate_slice(sphere3, anchor = anchor3, eps = 0.005, pch_other=26)
```

# Examples
XXX choose 6

## Geometric shapes
### Hollow sphere
We sample points on the surface of a sphere with radius $r=1$ in $p=3$ and $p=5$ dimensions. Using the `geozoo` package [@geozoo], we draw 2000 (5000) points that are uniformly distributed on the surface of a 3d (5d) sphere.
For both examples we start by slicing through the origin with the default parameters, in particular $\epsilon = 0.001$, i.e. $h = 0.1$ in 3d and $h = 0.25$ in 5d. Slicing through the origin results in sections where the points inside the slice are approximately on a circle with the full radius, and each view is similar to that shown in Figure~\ref{fig:renderFunction}. The animated gifs are available at XXX.

It is especially instructive to look at the sliced spheres with an off-center anchor point. In this case the different views obtained in the slice tour reveal more of the concave structure of the distribution. Figure \ref{fig:3anchored} shows example views from a slice tour on a 3d sphere with $r=1$ and anchor point $(0.7, 0.7, 0.7)$. Notice that the anchor point was chosen to fall outside the sphere. Depending on the viewing angle selected by the tour the slice contains points on the circle with radius $r$ (when the anchor point has a negligible orthogonal component to the viewing plane); a circle with radius $<r$ as the viewing angle is tilted away from the axis connecting the anchor point to the origin; a full circle with small radius as the angle increases; and finally we see an empty slice when the projection plane is orthogonal to the anchor point axis. A similar picture is found for the 5d sphere. Note however that as dimensionality increases, the larger orthogonal space means that the anchor point will have a large orthogonal component in most views.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.2\textwidth]{gifs/png/sphere-3-anchored-033.png}
\hspace{5mm}
\includegraphics[width=0.2\textwidth]{gifs/png/sphere-3-anchored-036.png}
\hspace{5mm}
\includegraphics[width=0.2\textwidth]{gifs/png/sphere-3-anchored-041.png}}
\caption{Different slices through a 3d hollow sphere with $r=1$, with anchor point $(0.7, 0.7, 0.7)$, showing full circle with small radius (left), circle with radius $<r$ (middle) and circle with radius $r$ (right).}
\label{fig:3anchored}
\end{figure*}

### Other geometric shapes

Examples: 4d cube, 4d torus

## Other examples

### Needle in a haystack

### Non-linear boundaries

# Summary and Outlook

In this note we introduce a new visualization method for systematic slicing of high-dimensional parameter spaces, based on interpolated projections obtained in a grand tour. The application to the hollow sphere demonstrates the potential of this new display to find and explore concave structures in distributions.

In applications with large number of dimensions and where only a few samples are available the method is limited. In this case thick slices may be needed to get a sufficient number of observations inside the slice, and as a result  small scale concavities cannot be resolved in the slice tour. If on the other hand a large number of samples are available one may consider displaying only points inside the slice to keep a clear display.

Instead of slicing in the full $p-d=p-2$ dimensional orthogonal space we can also use a subspace in which to slice. As pointed out in @prosection combining projections with sectioning in varying numbers of dimensions can reveal the codimension of an unknown distribution. One approach would be to interpolate between projection matrices with more than two columns, and use the first two for the display projection. The remaining directions then provide an orthogonal subspace that can be used for sectioning. A subtlety to consider is that geodesic interpolation is not ideal for such applications, as it ignores rotations between the directions in the projection matrix. As alternative interpolation methods are not currently available in the R implementation, we leave such extensions for future work.

In addition slice definitions beyond a simple cutoff on the orthogonal distance should be considered, where different definition can be used to highlight certain aspects of the data. We also expect interesting new developments by combining the slice tour and the guided tour into a weighted guided tour with index functions using the distance vector information in addition to the projected data points.
